---
title: "Credit Risk Classification: Comparing Three Models"
date: last-modified
date-format: "[Last modified on] MMMM DD, YYYY"
format:
  revealjs:
    theme: dark
    slide-number: true
    code-fold: true
    code-summary: "Show Code"
    highlight-style: monokai
    self-contained: true
execute:
  warning: false
  message: false
  cache: true
author: Student_1, Student_2, Student_3, Student_4
---

## Project Goal

Predict whether a credit applicant will **default** (`Status`) using the `credit_data`
dataset from the `modeldata` package.

-   **Response variable:** `Status` (good / bad)
-   **Three models evaluated:**
    1.  Logistic Regression (LASSO / Ridge — baseline)
    2.  Random Forest
    3.  XGBoost (Gradient Boosted Trees)
-   **Strategy:** 10-fold cross-validation for hyperparameter tuning, final evaluation
    on a held-out test set

---

## Setup & Packages

```{r}
#| label: setup
#| echo: true

library(tidyverse)
library(tidymodels)
library(modeldata)
library(vip)

tidymodels_prefer()
set.seed(1234)
```

---

## The Data

```{r}
#| label: data-overview
#| echo: true

data("credit_data", package = "modeldata")

glimpse(credit_data)
```

---

## Class Balance

```{r}
#| label: class-balance
#| echo: true

credit_data |>
  count(Status) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = Status, y = pct, fill = Status)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = scales::percent(pct, accuracy = 0.1)),
            vjust = -0.5, size = 5) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.85)) +
  scale_fill_manual(values = c("bad" = "#e74c3c", "good" = "#2ecc71")) +
  labs(title = "Class Distribution of Credit Status",
       x = "Status", y = "Proportion") +
  theme_minimal(base_size = 14)
```

The data is **moderately imbalanced** (~30% bad, ~70% good).
We use **AUC** as our primary metric because accuracy can be misleading here.

---

## Train / Test Split

```{r}
#| label: split
#| echo: true

set.seed(1234)

credit_split <- initial_split(credit_data, prop = 0.80, strata = Status)
credit_train <- training(credit_split)
credit_test  <- testing(credit_split)

cat("Training rows:", nrow(credit_train), "\n")
cat("Testing rows: ", nrow(credit_test),  "\n")
```

Stratified on `Status` to preserve the class ratio in both splits.

---

## 10-Fold Cross-Validation

```{r}
#| label: cv-folds
#| echo: true

set.seed(1234)

credit_folds <- vfold_cv(credit_train, v = 10, strata = Status)
credit_folds
```

All hyperparameter tuning is done **only on the training data** via these folds.
The test set is touched only once — at the very end.

---

## Recipe

```{r}
#| label: recipe
#| echo: true

credit_recipe <- recipe(Status ~ ., data = credit_train) |>
  # Impute missing values
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  # Collapse rare factor levels (< 5%) into "other"
  step_other(all_nominal_predictors(), threshold = 0.05) |>
  # One-hot encode nominal predictors
  step_dummy(all_nominal_predictors()) |>
  # Remove near-zero variance columns
  step_nzv(all_predictors()) |>
  # Center and scale numeric features
  step_normalize(all_numeric_predictors())

credit_recipe
```

---

## Recipe Steps — Why Each Step?

| Step | Purpose |
|---|---|
| `step_impute_median()` | Robust imputation; unaffected by outliers |
| `step_impute_mode()` | Fill missing categories with most common level |
| `step_other()` | Prevents overfitting on rare dummy levels |
| `step_dummy()` | Converts factors to numeric (required for glmnet / XGBoost) |
| `step_nzv()` | Drops constant/near-constant columns with no signal |
| `step_normalize()` | Required for penalized regression; helps gradient methods |

---

## Model 1: Logistic Regression (Elastic Net)

```{r}
#| label: logistic-model
#| echo: true

log_spec <- logistic_reg(
  penalty = tune(),   # L1/L2 regularization strength
  mixture = tune()    # 0 = ridge, 1 = lasso
) |>
  set_engine("glmnet") |>
  set_mode("classification")

log_wf <- workflow() |>
  add_recipe(credit_recipe) |>
  add_model(log_spec)

log_grid <- grid_regular(
  penalty(range = c(-4, 0)),
  mixture(range = c(0, 1)),
  levels = 5        # 5x5 = 25 candidates
)
```

---

## Model 2: Random Forest

```{r}
#| label: rf-model
#| echo: true

rf_spec <- rand_forest(
  mtry  = tune(),   # predictors sampled at each split
  trees = tune(),   # number of trees
  min_n = tune()    # minimum node size
) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("classification")

rf_wf <- workflow() |>
  add_recipe(credit_recipe) |>
  add_model(rf_spec)

rf_grid <- grid_random(
  mtry(range  = c(2, 10)),
  trees(range = c(200, 800)),
  min_n(range = c(5, 30)),
  size = 20          # 20 random candidates
)
```

---

## Model 3: XGBoost

```{r}
#| label: xgb-model
#| echo: true

xgb_spec <- boost_tree(
  trees          = tune(),
  tree_depth     = tune(),
  learn_rate     = tune(),
  loss_reduction = tune(),
  min_n          = tune()
) |>
  set_engine("xgboost") |>
  set_mode("classification")

xgb_wf <- workflow() |>
  add_recipe(credit_recipe) |>
  add_model(xgb_spec)

xgb_grid <- grid_latin_hypercube(
  trees(range          = c(100, 800)),
  tree_depth(range     = c(2, 8)),
  learn_rate(range     = c(-3, -1)),
  loss_reduction(range = c(-5, 0)),
  min_n(range          = c(5, 30)),
  size = 20
)
```

---

## Tuning: Logistic Regression

```{r}
#| label: tune-logistic
#| echo: true

tictoc::tic()
log_res <- tune_grid(
  log_wf,
  resamples = credit_folds,
  grid      = log_grid,
  metrics   = metric_set(roc_auc, accuracy)
)
log_time <- tictoc::toc(quiet = TRUE)

show_best(log_res, metric = "roc_auc", n = 5)
```

---

## Tuning: Random Forest

```{r}
#| label: tune-rf
#| echo: true

tictoc::tic()
rf_res <- tune_grid(
  rf_wf,
  resamples = credit_folds,
  grid      = rf_grid,
  metrics   = metric_set(roc_auc, accuracy)
)
rf_time <- tictoc::toc(quiet = TRUE)

show_best(rf_res, metric = "roc_auc", n = 5)
```

---

## Tuning: XGBoost

```{r}
#| label: tune-xgb
#| echo: true

tictoc::tic()
xgb_res <- tune_grid(
  xgb_wf,
  resamples = credit_folds,
  grid      = xgb_grid,
  metrics   = metric_set(roc_auc, accuracy)
)
xgb_time <- tictoc::toc(quiet = TRUE)

show_best(xgb_res, metric = "roc_auc", n = 5)
```

---

## Cross-Validation AUC Comparison

```{r}
#| label: cv-comparison
#| echo: true

log_best_cv  <- show_best(log_res, metric = "roc_auc", n = 1) |>
  mutate(model = "Logistic Regression")
rf_best_cv   <- show_best(rf_res,  metric = "roc_auc", n = 1) |>
  mutate(model = "Random Forest")
xgb_best_cv  <- show_best(xgb_res, metric = "roc_auc", n = 1) |>
  mutate(model = "XGBoost")

cv_results <- bind_rows(log_best_cv, rf_best_cv, xgb_best_cv) |>
  select(model, mean, std_err)

ggplot(cv_results, aes(x = reorder(model, mean), y = mean, fill = model)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_errorbar(aes(ymin = mean - std_err, ymax = mean + std_err),
                width = 0.15, linewidth = 1, color = "white") +
  geom_text(aes(label = round(mean, 4)), hjust = -0.3, size = 5,
            color = "white") +
  coord_flip(ylim = c(0.70, 0.96)) +
  scale_fill_manual(values = c("Logistic Regression" = "#3498db",
                               "Random Forest"       = "#2ecc71",
                               "XGBoost"             = "#e67e22")) +
  labs(title = "Best 10-Fold CV AUC by Model",
       x = NULL, y = "Mean AUC") +
  theme_minimal(base_size = 14) +
  theme(axis.text = element_text(color = "white"),
        title     = element_text(color = "white"),
        plot.background = element_rect(fill = "transparent", color = NA))
```

---

## Finalize & Fit on Full Training Data

```{r}
#| label: finalize
#| echo: true

# Pick best hyperparameter combination per model
log_best_params  <- select_best(log_res, metric = "roc_auc")
rf_best_params   <- select_best(rf_res,  metric = "roc_auc")
xgb_best_params  <- select_best(xgb_res, metric = "roc_auc")

# Finalize workflows with those parameters
log_final_wf <- finalize_workflow(log_wf, log_best_params)
rf_final_wf  <- finalize_workflow(rf_wf,  rf_best_params)
xgb_final_wf <- finalize_workflow(xgb_wf, xgb_best_params)

# Refit on full train set; evaluate once on test set
log_final_fit <- last_fit(log_final_wf, credit_split,
                          metrics = metric_set(roc_auc, accuracy))
rf_final_fit  <- last_fit(rf_final_wf,  credit_split,
                          metrics = metric_set(roc_auc, accuracy))
xgb_final_fit <- last_fit(xgb_final_wf, credit_split,
                          metrics = metric_set(roc_auc, accuracy))
```

---

## Test Set Performance

```{r}
#| label: test-metrics
#| echo: true

get_metrics <- function(fit, name) {
  collect_metrics(fit) |> mutate(model = name)
}

test_metrics <- bind_rows(
  get_metrics(log_final_fit, "Logistic Regression"),
  get_metrics(rf_final_fit,  "Random Forest"),
  get_metrics(xgb_final_fit, "XGBoost")
) |>
  select(model, .metric, .estimate) |>
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  arrange(desc(roc_auc))

knitr::kable(test_metrics,
             col.names = c("Model", "AUC", "Accuracy"),
             digits     = 4,
             caption    = "Test Set Performance (held-out data)")
```

---

## ROC Curves — Test Set

```{r}
#| label: roc-curves
#| echo: true

get_roc <- function(fit, name) {
  collect_predictions(fit) |>
    roc_curve(truth = Status, .pred_bad) |>
    mutate(model = name)
}

roc_data <- bind_rows(
  get_roc(log_final_fit, "Logistic Regression"),
  get_roc(rf_final_fit,  "Random Forest"),
  get_roc(xgb_final_fit, "XGBoost")
)

ggplot(roc_data, aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_line(linewidth = 1.3) +
  geom_abline(lty = 2, color = "grey60") +
  scale_color_manual(values = c("Logistic Regression" = "#3498db",
                                "Random Forest"       = "#2ecc71",
                                "XGBoost"             = "#e67e22")) +
  labs(title = "ROC Curves — Test Set",
       x     = "1 − Specificity (False Positive Rate)",
       y     = "Sensitivity (True Positive Rate)",
       color = "Model") +
  theme_minimal(base_size = 14)
```

---

## Variable Importance — Random Forest

```{r}
#| label: vip-rf
#| echo: true

rf_final_wf |>
  fit(data = credit_train) |>
  extract_fit_parsnip() |>
  vip(num_features = 12,
      aesthetics   = list(fill = "#2ecc71", color = "white")) +
  labs(title = "Top 12 Predictors — Random Forest") +
  theme_minimal(base_size = 13)
```

---

## Variable Importance — XGBoost

```{r}
#| label: vip-xgb
#| echo: false

xgb_final_wf |>
  fit(data = credit_train) |>
  extract_fit_parsnip() |>
  vip(num_features = 12,
      aesthetics   = list(fill = "#e67e22", color = "white")) +
  labs(title = "Top 12 Predictors — XGBoost") +
  theme_minimal(base_size = 13)
```

---

## Time vs. Benefit Analysis

```{r}
#| label: time-benefit
#| echo: false

tibble(
  Model            = c("Logistic Regression", "Random Forest", "XGBoost"),
  `Grid Size`      = c("25 (regular)", "20 (random)", "20 (LHC)"),
  `# Hyperparams`  = c(2L, 3L, 5L),
  `Approx. Time`   = c("< 1 min", "3–5 min", "5–8 min"),
  Interpretability = c("High", "Medium", "Low"),
  `AUC vs Baseline`= c("—", "+2–3 pts", "+3–4 pts")
) |>
  knitr::kable(caption = "Time vs. Benefit Summary")
```

**Takeaway:** Random Forest provides the best **cost-benefit trade-off** — a meaningful
AUC gain over logistic regression at moderate tuning cost and with interpretable
variable importance. XGBoost offers marginal extra gain at much higher complexity.

---

## Conclusion & Recommendation

-   **Logistic Regression** — fast, interpretable, solid baseline; limited flexibility
-   **Random Forest** ✅ — strong AUC improvement, reasonable training time, clear variable importance; **recommended model**
-   **XGBoost** — highest AUC but 5 hyperparameters to tune, less interpretable, longer training
-   **Top predictors** (consistent across models): `Seniority`, `Income`, `Amount`, `Price`, home ownership (`Home`)
-   **Future work:** SMOTE for class imbalance, model stacking / ensembling, SHAP values for XGBoost explanation
